From Light to GDP: A Study of Night-Time Light as an Indicator of Economic Development

### 2024-12-03 paper/paper.qmd
fixed some citations.

Economic growth measurement is central to policymaking, research, and international comparisons. Despite its importance, GDP figures, widely used as a benchmark for economic activity, are often criticized for inaccuracies, particularly in contexts where governance systems are less transparent. In authoritarian regimes, GDP data is vulnerable to manipulation for political ends, while in low-capacity states, weak statistical systems limit the reliability of national accounts. These challenges motivate the search for alternative indicators of economic activity. Night-Time Light (NTL) intensity, derived from satellite imagery, offers a compelling solution. Unlike GDP, NTL data is immune to tampering and provides a direct measure of human and industrial activity. By capturing emissions from streetlights, industrial areas, and residential zones, NTL reflects patterns of development that cannot be easily concealed or fabricated. This robustness, coupled with its availability across all regions, makes NTL an invaluable proxy for economic analysis.

Numerous studies have explored the relationship between NTL and GDP, particularly during the 1990s and early 2000s. These earlier studies primarily relied on the Defense Meteorological Satellite Program (DMSP) dataset, which was not originally designed for research purposes and suffered from technical limitations. More recent advancements, however, have introduced the Visible Infrared Imaging Radiometer Suite (VIIRS) dataset, which offers significantly improved quality, including better spatial resolution, radiometric calibration, and fewer issues with data saturation. With this enhanced dataset, spanning a broader temporal range and covering a larger sample of countries, it is now possible to revisit and expand upon earlier findings. This explore how the correlation between NTL and GDP has evolved, with particular attention to the moderating role of statistical capacity in different countries. Using fixed effects models, we aim to study the strength of the NTL-GDP correlation and assess its variation across countries grouped by statistical capacity.

@@

The structure of the paper is as follows. Section @sec-data outlines the data sources, the methods used to process and clean raster data from TIFF files, and the integration with World Bank data, along with a summary of the dataset. Section @sec-vari examines the distribution of key variables. Section @sec-rel visualizes and discusses the relationships between night-time light (NTL) and the variables of interest. Section @sec-model presents the different models used in the analysis, while Section @sec-results summarizes and interprets the model results. @sec-vari discusses the implciations of the findings and provides a broader discussion on uses and challenges of NTL. Finally, Section @sec-app includes supplementary figures and tables, such as the correlation matrix, model diagnostics, and an extended discussion on the measurement of night-time light.

@@

To associate each observation with a specific country, the extracted data were merged with a global shapefile of country boundaries through spatial join using [@geopandas]. Each latitude-longitude pair was assigned to the country in which it was located, and observations without an assigned country—primarily from international waters—were removed. Subsequently, to create an aggregate annual measure of NTL intensity for each country, all observations within a country were summed to calculate the total light emitted annually.

@@

```
In [@fig-dn-gdp-grade], we observe how this relationship between DN and GDP varies across countries with different statistical capacities. A list of countries categorized by their statistical capacity grades is given in [@fig-grades-cuntry] of the appendix. Countries with higher statistical capacity (graded A and B) exhibit a more tightly clustered and linear relationship, indicating consistent and robust data collection processes. Conversely, countries with lower statistical capacity (graded D and F) show more scatter and variability, suggesting potential discrepancies or weaker data reliability in their national accounts. This shows the intuitive idea that the strength of the DN-GDP relationship can serve as an indicator of the reliability of a country's national statistical systems.

@@

\vspace{2cm}
Building on the first model, the second model investigates how the relationship between NTL and GDP varies across countries with differing statistical capacities. As seen in Figure 1, the scatterplot of GDP and NTL faceted by statistical capacity suggests a weaker correlation in countries with lower statistical capacity. To explore this further, we include an interaction term between GDP and statistical capacity grade, allowing us to examine whether countries with stronger statistical systems exhibit a more consistent and robust relationship between NTL and GDP. The intuition is that higher statistical capacity is associated with better data quality and more stable economic structures, which may enhance the reliability of NTL as a proxy for economic activity. Similar to the first model, we control for year-specific and country-specific fixed effects to account for temporal and cross-country heterogeneity.
\vspace{1cm}
$$
\log(DN_{i,t}) = \beta_0 + \beta_1 \log(GDP_{i,t}) + \beta_2 (\text{Grade}_{i}) + \beta_3 (\log(GDP_{i,t}) \times \text{Grade}_{i}) + \gamma_t + \epsilon_{i,t}
$$

@@

In the first model, we examine the relationship between Night-Time Light (NTL) intensity and GDP, controlling for country and year-specific fixed effects. The coefficient for log(GDP) is positive and significant (0.974), indicating that a 1% increase in GDP corresponds to an approximately 0.97% increase in NTL intensity on average. The R² value of 0.958 suggests that the model explains 95.8% of the variation in NTL, highlighting a strong and robust relationship. The inclusion of fixed effects ensures that unobserved heterogeneity, such as geographic or temporal factors, is accounted for.

@@

Despite its numerous advantages, the use of NTL data faces significant challenges. A primary limitation is the temporal inconsistency and calibration issues associated with DMSP-OLS data. The lack of onboard calibration and sensor degradation introduces noise, complicating year-to-year comparisons. Furthermore, DMSP-OLS data suffer from saturation effects in highly illuminated urban areas and blooming effects in suburban and rural regions, leading to under- or overestimation of activity in these zones. The newer VIIRS dataset mitigates some of these issues with higher resolution and improved calibration but lacks the long historical coverage of DMSP-OLS, making it challenging to study trends over extended periods. @zhao2019 Additionally, variations in lighting intensity are influenced by factors unrelated to economic activity, such as atmospheric conditions, seasonal variations, and cultural practices, which can distort analyses. Another critical limitation is the inability to distinguish between different light sources, such as residential, industrial, or recreational lighting, reducing the specificity of NTL as a proxy for economic activities. Lastly, while combining NTL with ancillary datasets, such as NDVI or land surface temperature, improves its accuracy, it requires significant computational resources and robust modeling techniques. Addressing these limitations through improved calibration methods, cross-sensor integration, and the development of new algorithms is essential for maximizing the utility of NTL data in future research.

@@

Another avenue for future exploration lies in understanding how other variables, such as energy efficiency improvements, lighting technologies, and policy changes, influence the NTL-GDP relationship over time. The role of urbanization and industrialization, both key drivers of NTL, also warrants a more detailed examination, particularly in rapidly developing regions where structural changes are likely to alter the nature of this correlation. Incorporating spatial data and finer-resolution imagery could further refine our understanding, enabling a more localized assessment of economic activity.
\newpage
# Appendix {#sec-app}

@@

Advancements in satellite technology, such as onboard calibration systems and higher radiometric resolution, are reducing traditional limitations of NTL datasets. Modern sensors like VIIRS provide finer spatial and temporal resolutions, enabling more accurate monitoring of human activities. Integration of NTL data with other geospatial datasets, such as socio-demographic statistics, enhances its utility for interdisciplinary research. Future research could focus on developing methods to harmonize long-term datasets from multiple sensors, addressing issues like temporal inconsistency and cross-sensor calibration. Additionally, exploring machine learning techniques for extracting patterns from NTL data could offer new insights into urbanization, development, and disaster response. These advancements hold the promise of making NTL data even more robust and reliable for understanding global socioeconomic dynamics.

>>>>

Economic growth measurement is central to policymaking, research, and international comparisons. Despite its importance, GDP figures, widely used as a benchmark for economic activity, are often criticized for inaccuracies, particularly in contexts where governance systems are less transparent. In authoritarian regimes, GDP data is vulnerable to manipulation for political ends, while in low-capacity states, weak statistical systems limit the reliability of national accounts. These challenges motivate the search for alternative indicators of economic activity. Night-Time Light (NTL) intensity, derived from satellite imagery, offers a compelling solution. Unlike GDP, NTL data is immune to tampering and provides a direct measure of human and industrial activity. By capturing emissions from streetlights, industrial areas, and residential zones, NTL reflects patterns of development that cannot be easily concealed or fabricated.

Numerous studies have explored the relationship between NTL and GDP, particularly during the 1990s and early 2000s. These earlier studies primarily relied on the Defense Meteorological Satellite Program (DMSP) dataset, which was not originally designed for research purposes and suffered from technical limitations. More recent advancements, however, have introduced the Visible Infrared Imaging Radiometer Suite (VIIRS) dataset, which offers significantly improved quality, including better spatial resolution, radiometric calibration, and fewer issues with data saturation. With this enhanced dataset, spanning a broader temporal range and covering a larger sample of countries, it is now possible to revisit and expand upon earlier findings. This explore how the correlation between NTL and GDP has evolved, with particular attention to the moderating role of statistical capacity in different countries. Using fixed effects models, we aim to study the strength of the NTL-GDP correlation and assess its variation across countries grouped by statistical capacity. In particular, we aim to estimate the elasticity of GDP with respect to NTL intensity, capturing how changes in light emissions correspond to economic output, while controlling for country and year effects.

@@

The structure of the paper is as follows. @sec-data outlines the data sources, the methods used to process and clean raster data from TIFF files, and the integration with World Bank data, along with a summary of the dataset. @sec-vari examines the distribution of key variables. @sec-rel visualizes and discusses the relationships between night-time light (NTL) and the variables of interest.@sec-model presents the different models used in the analysis, while @sec-results summarizes and interprets the model results. @sec-vari discusses the implciations of the findings and provides a broader discussion on uses and challenges of NTL. Finally, the appendix - [@sec-last] includes supplementary figures and tables, such as the correlation matrix, model diagnostics, and an extended discussion on the measurement of night-time light.

@@

To associate each observation with a specific country, the extracted data were merged with a global shapefile of country boundaries through spatial join using 'geopandas' [@geopandas]. Each latitude-longitude pair was assigned to the country in which it was located, and observations without an assigned country—primarily from international waters—were removed. Subsequently, to create an aggregate annual measure of NTL intensity for each country, all observations within a country were summed to calculate the total light emitted annually.

@@

```
In [@fig-dn-gdp-grade], we observe how this relationship between DN and GDP varies across countries with different statistical capacities. A list of countries categorized by their statistical capacity grades is given in [@fig-grades-cuntry] of the appendix. Countries with higher statistical capacity (graded A and B) exhibit a more tightly clustered and linear relationship. Conversely, countries with lower statistical capacity (graded D and F) show more scatter and variability, suggesting potential discrepancies or weaker data reliability in their national accounts. This shows the intuitive idea that the strength of the DN-GDP relationship can serve as an indicator of the reliability of a country's national statistical systems.

@@

\vspace{2cm}
Building on the first model, the second model investigates how the relationship between NTL and GDP varies across countries with differing statistical capacities. As seen in Figure 1, the scatterplot of GDP and NTL faceted by statistical capacity suggests a weaker correlation in countries with lower statistical capacity. To explore this further, we include an interaction term between GDP and statistical capacity grade, allowing us to examine whether countries with stronger statistical systems exhibit a more consistent relationship between NTL and GDP. The intuition is that higher statistical capacity is associated with better data quality and more stable economic structures, which may enhance the reliability of NTL as a proxy for economic activity. Similar to the first model, we control for year-specific and country-specific fixed effects to account for temporal and cross-country heterogeneity.
\vspace{1cm}
$$
\log(DN_{i,t}) = \beta_0 + \beta_1 \log(GDP_{i,t}) + \beta_2 (\text{Grade}_{i}) + \beta_3 (\log(GDP_{i,t}) \times \text{Grade}_{i}) + \gamma_t + \epsilon_{i,t}
$$

@@

In the first model, we examine the relationship between Night-Time Light (NTL) intensity and GDP, controlling for country and year-specific fixed effects. The coefficient for log(GDP) is positive and significant (0.974), indicating that a 1% increase in GDP corresponds to an approximately 0.97% increase in NTL intensity on average. The R² value of 0.958 suggests that the model explains 95.8% of the variation in NTL, highlighting a strong relationship. The inclusion of fixed effects ensures that unobserved heterogeneity, such as geographic or temporal factors, is accounted for.

@@

Despite its numerous advantages, the use of NTL data faces significant challenges. A primary limitation is the temporal inconsistency and calibration issues associated with DMSP-OLS data. The lack of onboard calibration and sensor degradation introduces noise, complicating year-to-year comparisons. Furthermore, DMSP-OLS data suffer from saturation effects in highly illuminated urban areas and blooming effects in suburban and rural regions, leading to under- or overestimation of activity in these zones. The newer VIIRS dataset mitigates some of these issues with higher resolution and improved calibration but lacks the long historical coverage of DMSP-OLS, making it challenging to study trends over extended periods. @Zhao2019 Additionally, variations in lighting intensity are influenced by factors unrelated to economic activity, such as atmospheric conditions, seasonal variations, and cultural practices, which can distort analyses. Another critical limitation is the inability to distinguish between different light sources, such as residential, industrial, or recreational lighting, reducing the specificity of NTL as a proxy for economic activities. Lastly, while combining NTL with ancillary datasets, such as NDVI or land surface temperature, improves its accuracy, it requires significant computational resources and modeling techniques. Addressing these limitations through improved calibration methods, cross-sensor integration, and the development of new algorithms is essential for maximizing the utility of NTL data in future research.

@@

Another avenue for future exploration lies in understanding how other variables, such as energy efficiency improvements, lighting technologies, and policy changes, influence the NTL-GDP relationship over time. The role of urbanization and industrialization, both key drivers of NTL, also warrants a more detailed examination, particularly in rapidly developing regions where structural changes are likely to alter the nature of this correlation. Incorporating spatial data and finer-resolution imagery could further refine our understanding, enabling a more localized assessment of economic activity.
\newpage

# Appendix {#sec-last}

@@

Advancements in satellite technology, such as onboard calibration systems and higher radiometric resolution, are reducing traditional limitations of NTL datasets. Modern sensors like VIIRS provide finer spatial and temporal resolutions, enabling more accurate monitoring of human activities. Integration of NTL data with other geospatial datasets, such as socio-demographic statistics, enhances its utility for interdisciplinary research. Future research could focus on developing methods to harmonize long-term datasets from multiple sensors, addressing issues like temporal inconsistency and cross-sensor calibration. Additionally, exploring machine learning techniques for extracting patterns from NTL data could offer new insights into urbanization, development, and disaster response. These advancements hold the promise of making NTL data even more reliable for understanding global socioeconomic dynamics.
